{\vspace{-2cm}}Temporal Congestion Analysis and Predictive Modeling of Ride-Hailing Trips in NYC Using High-Volume FHV Data

### 2025-04-22 paper.tex
chore: Draft paper.

\section{Background Review}
Recent work by Riley et al. (2020) proposes an integrated optimization and machine learning framework to improve real-time ride sharing dispatch using NYC trip data, reducing average passenger wait times by more than 50\% \cite{riley2020dispatching}. Similarly, Ma et al. introduce a congestion-aware dispatch and charging scheduling system for electric ride-hailing fleets, achieving improved profitability and service efficiency in high-congestion scenarios \cite{ma2024congestionaware}. These studies highlight advanced data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.

@@

Several works have used this data set to assess the impact of ride-hailing on traffic flow and use of public transit. For example, a study by the NYU Tandon School of Engineering found that the New York City congestion surcharge reduced the number of ride-hailing trips, but did not significantly alleviate congestion \cite{nyu2023}. Furthermore, research from the CUNY Graduate Center examined the effects of the 2019 congestion surcharge on FHV usage in Manhattan, showing a decline in rides in charged zones \cite{cuny2023}.

@@

\subsection{Data Processing}
Data preprocessing plays a critical role in reducing both storage and compute requirements, resulting in not only time savings but also significant cost savings in a cloud-based environment. It also reduces the cognitive load on users by minimizing the number of columns they need to interpret and provides a standardized clean dataset that users can reliably use as the foundation for further analysis. By filtering out invalid records and outliers early in the pipeline, I can minimize the volume of data that needs to be processed in downstream stages, leading to more efficient use of Spark resources.

@@

\subsection{Operationalization and Scalability}
To transition from a one-off analysis to a production-ready system, I propose extending the workflow to support automation and real-time data ingestion. Automated preprocessing of monthly datasets can be achieved by designing a parameterized PySpark pipeline and scheduling it with tools such as Apache Airflow \cite{airflow2024}. This setup would automatically clean and transform the dataset for each new month after publication, improving efficiency and standardization.

In addition, the architecture could be extended to support real-time ingestion and streaming analytics. By integrating tools such as Apache Kafka with Spark Structured Streaming \cite{structuredstreaming2024}, the system could handle continuous data feeds from ride-hailing APIs or city sensors. This would enable use cases such as live congestion monitoring, real-time alerting, and dynamic traffic optimization. Operationalizing the pipeline in this way improves scalability and real-world applicability for ongoing urban analytics.

>>>>

\section{Background Review}
Riley et al. propose a scalable real-time ride-sharing architecture (A-RTRS) that integrates optimization, machine learning, and predictive model control for dynamic dispatching and vehicle relocation on idle. Their system uses time-batched optimization, historical demand forecasting, and zone-level repositioning to reduce average passenger wait times by up to 55\%. This architecture demonstrates how predictive and adaptive mechanisms can operate in large-scale, high-velocity transportation environments. \cite{riley2020dispatching}. Ma et al. present a congestion-aware optimization framework for electric ride-hailing fleets that jointly coordinates vehicle dispatching and charging operations. Their system includes a day-ahead charging plan based on time-of-use energy pricing and a reactive model that adapts to vehicle availability and charger congestion in real time. Evaluated using simulated NYC taxi data, their congestion-aware approach outperformed benchmark policies in both service rate and profit, demonstrating the impact of integrated vehicle and charging scheduling on operational efficiency. \cite{ma2024congestionaware}. These studies highlight advanced data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.

@@

Several works have used this data set to assess the impact of ride-hailing on traffic flow and use of public transit. A study by the NYU Tandon School of Engineering empirically evaluated the impact of NYC’s ride-hailing congestion surcharge policy using a difference-in-differences (DiD) approach. Although the policy led to a significant reduction in travel demand - 11\% overall, with Lyft experiencing the greatest drop - it did not result in a significant reduction in congestion. The study highlights that travel reductions were more pronounced for short-distance rides and in areas with alternative transportation options such as subways and bike sharing. It also found that the policy disproportionately affected lower income riders, raising concerns about equity and policy effectiveness\cite{nyu2023}. Although the Tandon study provides valuable insight into policy impacts using historical data, it lacks a supporting big data system architecture to enable real-time monitoring or operationalization. In contrast, the architecture proposed in this paper, which combines automated batch processing, real-time ingestion, and predictive modeling, could support dynamic congestion policy evaluation and feedback loops. Such a system could enable policy makers to monitor the impact of surcharges or routing incentives as they occur, enabling more responsive and equitable management of urban mobility. Furthermore, research from the CUNY Graduate Center examined the effects of the 2019 congestion surcharge on FHV usage in Manhattan, showing a decline in rides in charged zones \cite{cuny2023}. Although insightful, the CUNY study does not propose a scalable system for continuous or predictive policy evaluation. The big data architecture in this project would enable real-time pickup and collision monitoring, allowing faster feedback on congestion interventions and even potential automation of surcharge adjustments based on live trends. This infrastructure-oriented approach supports both exploratory and operational use cases that go beyond the economic evaluation described in the CUNY study.

@@

\subsection{Data Processing}
Data preprocessing plays a critical role in reducing both storage and compute requirements, resulting in not only time savings but also significant cost savings in a cloud-based environment. It also reduces the cognitive load on users by minimizing the number of columns they need to interpret and provides a standardized clean dataset that users can reliably use as the foundation for further analysis. By filtering out invalid records and outliers early in the pipeline, I can minimize the volume of data that needs to be processed in downstream stages, leading to more efficient use of Spark and storage resources.

@@

\subsection{Operationalization and Scalability}
To transition from a one-off analysis to a production-ready system, I propose extending the workflow to support automation and real-time data ingestion. Automated preprocessing of monthly datasets can be achieved by designing a parameterized PySpark pipeline and scheduling it with tools such as Apache Airflow \cite{airflow2024}. This setup would automatically clean and transform the dataset for each new month after publication, improving efficiency and standardization. Figure \ref{fig:big-data-architecture} outlines a potential production-level architecture to support batch ingestion, scalable processing, secure storage, and model analysis.

In addition, the architecture could be extended to support real-time ingestion and streaming analytics. By integrating tools such as Apache Kafka with Spark Structured Streaming \cite{structuredstreaming2024}, the system could handle continuous data feeds from ride-hailing APIs or city sensors. This would enable use cases such as live congestion monitoring, real-time alerting, and dynamic traffic optimization. Operationalizing the pipeline in this way improves scalability and real-world applicability for continuous analytics.

\begin{figure}
\includegraphics[width=\textwidth]{images/big-data-architecture.png}
\caption{Production-level architecture}
\label{fig:big-data-architecture}
\end{figure}
